Edge sign prediction in general graphs and connection with Correlation Clustering

### 2017-07-08 gtx_algorithm.tex
typos
found a mistake in the node reduction proof :(

\begin{algorithm}[tbh]
  \caption{\gtx{}($G_0=(V_0,E_0)$) \label{alg:gtx}}
	\begin{algorithmic}[1]
    \State Let $eccentricity$ be an array of size $|V_0|$ initially all set to $0$
    \Let{$G_t$}{$G_0$}
    \Let{$inner\_edges\_seq$, $outer\_edges\_seq$}{$[]$, $[]$}
    \Repeat
      \Let{$stars$, $inner\_edges$, $membership$}{\Call{Extract-Stars}{$G_t$}}
      \State Add $inner\_edges$ to the list $inner\_edges\_seq$
      \State \Call{Update-Eccentricity}{$stars$, $eccentricity$, $original\_node$}
      \Let{$full\_membership$, $original\_node$}{\Call{Update-Nodes-Mapping}{$full\_membership$, $membership$}}
      \Let{$G_{t+1}$, $\!outer\_edges$}{\Call{Collapse-Stars}{$E_t \!\setminus\! inner\_edges$,
      $\!membership$, $\!eccentricity$}}
      \State Add $outer\_edges$ to the list $outer\_edges\_seq$
      \Let{$G_{t}$}{$G_{t+1}$}
    \Until{$|outer\_edges|>0$}
    \State \textbf{return} \Call{Assemble-Spanning-Tree}{$inner\_edges\_seq$, $outer\_edges\_seq$}
		\begin{center}
      \vspace{-.5\baselineskip}
			\rule{0.5\textwidth}{.2pt}
		\end{center}
    \includegraphics{assets/tmp-code/update_eccentricity.pdf}
    \includegraphics{assets/tmp-code/update_nodes_mapping.pdf}
    \includegraphics{assets/tmp-code/assemble_spanning_tree.pdf}
	\end{algorithmic}
\end{algorithm}

@@

Such singleton stars are problematic because when too many are created, we can end up having
$|V_{t+1}|$ almost as large as $|V_t|$. Consider for instance the graph $G_t$ at the top of
\autoref{fig:gtx_manysingles} (which is actually a tree). It has of order $k^3$ nodes and because
all nodes at level 2 will be selected as center of their star, the graph $G_{t+1}$ (showed at the
bottom of \autoref{fig:gtx_manysingles}) also have around $k^3$ nodes, meaning that
$\frac{|V_{t+1}|}{|V_t|}$ tends to $1$ as $k$ goes to infinity. This is because
$k\left(k(k+1)\right)$ singleton stars were created. Let us denote the fraction of singleton stars
over the total number of nodes by $\alpha$. In that case,
$$\alpha = \frac{k\left(k(k+1)\right)}{1+k+k(k+1)+k^2(k+1)}
= 1 - \frac{1+2k+k^2}{1+2k+2k^2+k^3}
= 1 - \frac{1+k}{1+k+k^2}$$
We will now see that when so many singleton stars are created at one iteration, the next iteration
greatly reduces the number of nodes so that over the course of two iterations, we have the following
result:

\begin{lemma}
  \label{lemma:gtx_lessnodes}
  During the execution of the \gtx{} algorithm, for all $t \leq K-2$, $|V_{t+2}| \leq
  \frac{2}{3}|V_t|$.
\end{lemma}
\begin{marginfigure}
  \centering
  \includegraphics[width=\linewidth]{assets/raw/lots_of_singletons_bw.png}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_manysingles}
\end{marginfigure}
\begin{proof}
  Say that at iteration $t$ of \gtx{}, we have $n_t$ nodes and assume a fraction $\alpha$ of them
  will form singleton stars during the \tth{} iteration. This means there will be at most $\frac{1-\alpha}{2}n_t$
  \emph{proper} stars, because such stars are made of at least two nodes. Now fix a constant
  $\lambda \in (0,1)$, for instance $\lambda = \nicefrac{9}{10}$. If $\alpha \leq \lambda$, $n_{t+1}
  \leq (\frac{1-\alpha}{2} + \alpha)n_t = \frac{1+\alpha}{2}n_t$ and $\frac{1+\alpha}{2}n_t \leq
  \frac{1+\lambda}{2}n_t = \frac{19}{20}n_t$ so we get a
  constant rate reduction of the number of nodes. On the other hand, if $\alpha > \lambda$, we could
  be in the situation of the previous example and have $\alpha$ be arbitrarily close to $1$. Then we
  only have a few proper stars and by construction, each singleton star is connected to least one proper star.
  In the worst case, those connections are uniform so that each of these proper stars has degree at least
  $\frac{\alpha n_t}{\frac{1-\alpha}{2}n_t} = \frac{2\alpha}{1-\alpha}$ which goes to infinity when
  $\alpha$ goes to $1$ and is greater than $18$ when $\alpha$ is greater than $\nicefrac{9}{10}$.
  Even ignoring the fact that those proper stars are connected, it follows that in the next
  iteration, the number of node will be reduced by the degree of those proper stars. Setting in the
  worst case $n_{t+1}=\frac{1+\alpha}{2}n_t$, we thus have $$n_{t+2} \leq
  \frac{n_{t+1}}{\frac{2\alpha}{1-\alpha}} = \frac{1-\alpha^2}{4\alpha}n_t$$
  \begin{marginfigure}
    \centering
    \includegraphics[width=\linewidth]{assets/tmp-code/gtx_reduc.pdf}
    \caption{The fraction of $n_t$ nodes remaining in $n_{t+2}$ for the three cases described in the
    main text.}
    % \label{fig:gtx_manysingles}
  \end{marginfigure}
  This defines three ways to go from $n_t$ to $n_{t+2}$: if the first iteration has $\alpha>\lambda$,
  then we just seen that in the worst case we have at least $n_{t+2}\leq
  \frac{1-\lambda^2}{4\lambda}n_t$; if the first iteration has $\alpha < \lambda$, then either the
  second also has  $\alpha < \lambda$ (in which case $n_{t+2}\leq
  \left(\frac{1+\lambda}{2}\right)^2n_t$) or the second has $\alpha>\lambda$ (in which case
  $n_{t+2}\leq \frac{1+\lambda}{2}n_t$). We can optimize over $\lambda$ by finding its value such
  that $\frac{1-\lambda^2}{4\lambda}=\frac{1+\lambda}{2}$, yielding $\lambda = \frac{1}{3}$. Then,
  in any event, $n_{t+2}\leq \frac{2}{3}n_t$ as claimed.
\end{proof}

From that we conclude that the number of iterations of the \gtx{} algorithm satisfies $K=O(\log
|V_0|)$. We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed, we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.

>>>>

\begin{algorithm}[tbh]
  \caption{\gtx{}($G_0=(V_0,E_0)$) \label{alg:gtx}}
	\begin{algorithmic}[1]
    \State Let $eccentricity$ be an array of size $|V_0|$ initially all set to $0$
    \Let{$G_t$}{$G_0$}
    \Let{$inner\_edges\_seq$, $outer\_edges\_seq$}{$[]$, $[]$}
    \Repeat
      \Let{$stars$, $inner\_edges$, $star\_membership$}{\Call{Extract-Stars}{$G_t$}}
      \State Add $inner\_edges$ to the list $inner\_edges\_seq$
      \State \Call{Update-Eccentricity}{$stars$, $eccentricity$, $original\_node$}
      \Let{$full\_membership$, $original\_node$}{\Call{Update-Nodes-Mapping}{$full\_membership$,
      $star\_membership$}}
      \Let{$G_{t+1}$, $\!outer\_edges$}{\Call{Collapse-Stars}{$E_t \!\setminus\! inner\_edges$,
      $\!star\_membership$, $\!eccentricity$}}
      \State Add $outer\_edges$ to the list $outer\_edges\_seq$
      \Let{$G_{t}$}{$G_{t+1}$}
    \Until{$|outer\_edges|>0$}
    \State \textbf{return} \Call{Assemble-Spanning-Tree}{$inner\_edges\_seq$, $outer\_edges\_seq$}
		\begin{center}
      \vspace{-.5\baselineskip}
			\rule{0.5\textwidth}{.2pt}
		\end{center}
    \includegraphics{assets/tmp-code/update_eccentricity.pdf}
    \includegraphics{assets/tmp-code/update_nodes_mapping.pdf}
    \includegraphics{assets/tmp-code/assemble_spanning_tree.pdf}
	\end{algorithmic}
\end{algorithm}

@@

Such singleton stars are problematic because when too many are created, we can end up having
$|V_{t+1}|$ almost as large as $|V_t|$. Consider for instance the graph $G_t$ at the top of
\autoref{fig:gtx_manysingles} (which is actually a tree). It has of order $k^3$ nodes and because
all nodes at level 2 will be selected as center of their star, the graph $G_{t+1}$ (showed at the
bottom of \autoref{fig:gtx_manysingles}) also have around $k^3$ nodes, meaning that
$\frac{|V_{t+1}|}{|V_t|}$ tends to $1$ as $k$ goes to infinity. This is because
$k\left(k(k+1)\right)$ singleton stars were created. Let us denote the fraction of singleton stars
over the total number of nodes by $\alpha$. In that case,
$$\alpha = \frac{k\left(k(k+1)\right)}{1+k+k(k+1)+k^2(k+1)}
= 1 - \frac{1+2k+k^2}{1+2k+2k^2+k^3}
= 1 - \frac{1+k}{1+k+k^2}$$
We will now see that when so many singleton stars are created at one iteration, the next iteration
greatly reduces the number of nodes so that over the course of two iterations, we have the following
result:
{\color{blue}{\large the following is actually incorrect but I leave it for now for illustration
purpose}

\begin{lemma}
  \label{lemma:gtx_lessnodes}
  During the execution of the \gtx{} algorithm, for all $0\leq t \leq K-2$, $|V_{t+2}| \leq
  \frac{2}{3}|V_t|$.
\end{lemma}
\begin{marginfigure}
  \centering
  \includegraphics[width=\linewidth]{assets/raw/lots_of_singletons_bw.png}
  \caption{A case were too many singletons \enquote{waste} one iteration of \gtx{}}
  \label{fig:gtx_manysingles}
\end{marginfigure}
\begin{proof}
  Say that at iteration $t$ of \gtx{}, we have $n_t$ nodes and assume a fraction $\alpha$ of them
  will form singleton stars during the \tth{} iteration. This means there will be at most $\frac{1-\alpha}{2}n_t$
  \emph{proper} stars, because such stars are made of at least two nodes. Now fix a constant
  $\lambda \in (0,1)$, for instance $\lambda = \nicefrac{9}{10}$. If $\alpha \leq \lambda$, $n_{t+1}
  \leq (\frac{1-\alpha}{2} + \alpha)n_t = \frac{1+\alpha}{2}n_t$ and $\frac{1+\alpha}{2}n_t \leq
  \frac{1+\lambda}{2}n_t = \frac{19}{20}n_t$ so we get a
  constant rate reduction of the number of nodes. On the other hand, if $\alpha > \lambda$, we could
  be in the situation of the previous example and have $\alpha$ be arbitrarily close to $1$. Then we
  only have a few proper stars and by construction, each singleton star is connected to least one proper star.
  In the worst case, those connections are uniform so that each of these proper stars has degree at least
  $\frac{\alpha n_t}{\frac{1-\alpha}{2}n_t} = \frac{2\alpha}{1-\alpha}$ which goes to infinity when
  $\alpha$ goes to $1$ and is greater than $18$ when $\alpha$ is greater than $\nicefrac{9}{10}$.
  \textbf{Even ignoring the fact that those proper stars are connected}, it follows that in the next
  iteration, the number of node will be reduced by the degree of those proper stars. Setting in the
  worst case $n_{t+1}=\frac{1+\alpha}{2}n_t$, we thus have $$n_{t+2} \leq
  \frac{n_{t+1}}{\frac{2\alpha}{1-\alpha}} = \frac{1-\alpha^2}{4\alpha}n_t$$
  \begin{marginfigure}
    \centering
    \includegraphics[width=\linewidth]{assets/tmp-code/gtx_reduc.pdf}
    \caption{The fraction of $n_t$ nodes remaining in $n_{t+2}$ for the three cases described in the
    main text.}
    % \label{fig:gtx_manysingles}
  \end{marginfigure}
  This defines three ways to go from $n_t$ to $n_{t+2}$: if the first iteration has $\alpha>\lambda$,
  then we just seen that in the worst case we have at least $n_{t+2}\leq
  \frac{1-\lambda^2}{4\lambda}n_t$; if the first iteration has $\alpha < \lambda$, then either the
  second also has  $\alpha < \lambda$ (in which case $n_{t+2}\leq
  \left(\frac{1+\lambda}{2}\right)^2n_t$) or the second has $\alpha>\lambda$ (in which case
  $n_{t+2}\leq \frac{1+\lambda}{2}n_t$). We can optimize over $\lambda$ by finding its value such
  that $\frac{1-\lambda^2}{4\lambda}=\frac{1+\lambda}{2}$, yielding $\lambda = \frac{1}{3}$. Then,
  in any event, $n_{t+2}\leq \frac{2}{3}n_t$ as claimed.
\end{proof}

From that we conclude that the number of iterations of the \gtx{} algorithm satisfies $K=O(\log
|V_0|)$.}
The problem is that we cannot ignore the fact that proper stars are connected, since one of them can
be connected to all others and have the highest degree. That proper star is then chosen as the first
center, all other proper stars become its periphery and all their attached singletons star remain
singleton stars at the next iteration. This actually what happens on \autoref{fig:gtx_manysingles},
an example where $|V_{t+2}| \sim |V_t| \sim k^3$. This could even be worse if proper stars are
connected in such a hierarchical way that singleton stars can survive for many iterations.

We would like to prove a similar result on the edges so that the overall complexity of the
\gtx{} algorithm would be $O(|V_0|\log |V_0|+|E_0|)$. However, this does not follow directly from the
reduction of number of nodes. Indeed, we can have a situation where most of the edges of $E_t$ are
between the stars of $G_t$ and thus will remain in the next iteration. Still, we hope that
similarly to the previous proof, this high density of outer edges will lead to strong collapse in
the next iteration.
